---
title: "Day_08_scientific_visualization_mastery"
author: "By Muhammad Yasir Qurashi"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# **Differential Gene Expression**

## **Today's Goal**

The goal is to transform a gene expression matrix into a publication-quality Volcano plot & MAplot that highlights Statistical significance difference between genes in two different experimental conditions.

## Step 01a 

* Loading libraries & dataset

```{r}
library(DESeq2)
library(airway)
data("airway")

## DESeq2 package, which is one of the most widely used tools for differential gene expression analysis in RNA-seq studies.

## DESeq2 performs:

# 1. normalization

# 2. dispersion estimation

# 3. statistical testing; to determine which genes change significantly between conditions.
```

## Step 01b

* Extract count matrix & metadata

```{r}

counts <- assay(airway) # Extract gene expression count matrix
head(counts)

colData(airway) # Extract gene expression metadata

```

## Step 02

* Create DESeq dataset

```{r}

DE_data <- DESeqDataSet(airway, design = ~dex)

## DESeqDataSet creates Creates an object that stores: 1. Expression counts 2. Sample metadata 3. Experimental design; design tells DESeq to compare gene in different treatments
```


## Step 03 

* Filter low count reads

```{r}

DE_data <- DE_data[rowSums(counts(DE_data)) > 10, ]
                   
## RNA-seq datasets contain many genes with very low counts
 
# Low-count genes; 1. increase noise 2. reduce statistical power
```


## Step 04 

* Run Differential Expression analysis

```{r}

DE_data <- DESeq(DE_data)

```

DESeq2 performs three operations;

* Normalization

Adjusts for sequencing depth differences between samples

* Dispersion estimation

Measures biological variability for each gene

* Negative binomial model testing

Tests whether gene expression differs between conditions


## Step 05

* Extract significant results from DESeq 

```{r}
DE_result <- results(DE_data)
head(DE_result)
nrow(DE_result)
```

## Step 06

* Order gene by significance 

```{r}

DE_result <- DE_result[order(DE_result$padj), ]
head(DE_result)
top10_genes <- rownames(DE_result)[1:10]
```

## Step 07 

* Volcano plot 

```{r}
library(EnhancedVolcano)
# Converting DE_result to data frame

res_df <- as.data.frame(DE_result)
head(res_df)
class(res_df)
```

```{r, fig.dpi=1200}
EnhancedVolcano( res_df, x = 'log2FoldChange', y = 'padj', lab = row.names(res_df), pCutoff = 0.05, FCcutoff = 1, pointSize = 1, labSize = 2,
                 selectLab = top10_genes,
                 title = "Differential Expression: Dexamethasone Treatment",
                 col = c("grey30","forestgreen","royalblue","red2"),
                 subtitle = "Airway RNA-seq dataset",caption = "Made by Muhammad Yasir Qurashi", ylim = c(0,30), xlim = c(-5,+5)) +
  theme_bw()
```

* Innterpretation of VOlcano plot

| Color | Meaning                 |
| ----- | ----------------------- |
| grey  | not significant         |
| blue  | significant fold change |
| green | significant p-value     |
| red   | both significant        |


Differential expression analysis identified multiple genes significantly regulated by dexamethasone treatment. The volcano plot revealed a subset of genes with both large fold changes and strong statistical support (adjusted p-value < 0.05), indicating substantial transcriptional responses to glucocorticoid exposure.


Best Regards,

*Muhammad Yasir Qurashi*

Research Data Analysis Tools Mentor
