---
title: "Day_02_scientific_visualization_training"
author: "By Muhammad Yasir Qurashi"
date: "2026-03-02"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# One-way ANOVA in R

## Loading_libraries

```{r}
library(tidyverse)
library(ggpubr)
library(multcompView)
library(plotly)
library(RColorBrewer)
library(ggrain)
library(ggsci)
```

## Loading dataset

```{r}
TG <- ToothGrowth
View(TG)
summary(TG)
TG$dose <- as.factor(TG$dose)
summary(TG)
```

## RainCloud Plot

```{r, fig.width=7, fig.height=5}
p1 <- ggplot(data = TG, mapping = aes(x = dose, y = len, fill = dose)) +
  geom_rain(color = "blue")+
  ggtitle("Rain Cloud Plot")+
  theme_pubr()+
  scale_fill_bmj();p1
```

## Density plot

```{r,fig.width=7, fig.height=5}
# Basic density plot
p2 <- ggplot(TG, aes(x = len, fill = dose)) +
  geom_density(alpha = 0.6) +
  labs(
    title = "Density Plot",
    x = "Tooth Length",
    y = "Density"
  ) +
  theme_pubr() +
  theme(
    plot.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold")
  )+
  scale_fill_bmj();p2
```


## Hypothesis testing steps for one-way ANOVA

**1. Make your hypothesis**

null hypothesis --\> there is no significant difference in means of three groups

alternative --\> there is significant difference in mean of at least one group

**2. Define your significance level**

alpha = 0.05, C.I = 0.95

**3. Selection of test**

*One way ANOVA*

**4. Calculation**

```{r}
# Normality Check ( first condition of ANOVA)

TG %>% 
  group_by(dose) %>% 
  summarise(
  shapiro_p = shapiro.test(len)$p.value
  )

 # Data is normal
```

```{r}
# Homogenity ( Second Condition for ANOVA)
library(car)
leveneTest(len ~ dose, data = TG, alternative = "two.sided")
 # Data is Homogenous too
```

```{r}
# ANOVA calculation
results_anova <- aov(len ~ dose, data = TG)
summary(results_anova)
```

**5. Conclusion**

one way ANOVA has been done with statistical significant results ( n = 60, p_value \< 0.001 )

## Barplot without lettering

```{r, fig.width=7, fig.height=5}
p3 <- ggplot(data = TG, mapping = aes(x = dose, y = len, fill = dose))+
  geom_bar(stat = "summary", fun = "mean", width = 0.3, linewidth = 0.5)+
  geom_errorbar(stat = "summary", fun.data = mean_se, width = 0.05, linewidth = 0.7)+
  labs(x = "Dose", y = "Mean of Tooth Length", title = "Bar Plot", subtitle = "One way ANOVA without lettering", caption = "By Muhammad Yasir Qurashi")+
  stat_compare_means(method = "anova", label.y = 30) +
    theme(
    axis.title = element_text(size = 12, color = "black", face = "bold"),
    axis.text = element_text(size = 10, color = "black", face = "bold"), 
    axis.line = element_line(linewidth = 1, color = "black"),
    plot.title = element_text(size = 16, color = "blue", face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12,face = "bold"),
    legend.background = element_rect(color = "black", fill = "grey"),
    legend.position = "right",
    legend.text = element_text(size = 10, face = "bold"),
     panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank()
  ) +
  scale_fill_brewer(palette = "Dark2");p3
```

## Barplot with lettering

TO make barplot with significant letters there are four steps given below:

**1. Step 1**

Find group wise mean

```{r}
dose_mean <- TG %>% 
  group_by(dose) %>% 
  summarise(
    mean_len = mean(len),
    quan_len = quantile(len, 0.75)
  );dose_mean
```

**1. Step 2**

calculation of anova

```{r}
anova <- aov(len ~ dose, data = TG)
summary(anova)
```

**1. Step 3**

Post_hoc test

```{r}
tukey <- TukeyHSD(anova)
#View(tukey)
```

**1. Step 4**

Giving Significant Letters

```{r}
SL <-  multcompLetters4(anova,tukey)
#View(SL)

SL <- as.data.frame.list(SL$dose)
```

**1. Step 5**

Combine Significant letters with Mean Table

```{r}

dose_mean <- cbind(dose_mean, SL = SL$Letters)
#View(dose_mean)
```

# Graphical Representation

```{r, fig.width=8, fig.height=6}
p4 <- ggplot(data = TG, mapping = aes(x = dose, y = len, fill = dose))+
  geom_bar(stat = "summary", fun = "mean", width = 0.3, linewidth = 0.5)+
  geom_errorbar(stat = "summary", fun.data = mean_se, width = 0.05, linewidth = 0.7)+
  geom_text(data = dose_mean, aes(x = dose, y = mean_len, label = SL), size = 5, hjust = 0, vjust = -1.5)+
  labs(x = "Dose", y = "Mean of Tooth Length", title = "Bar Plot", subtitle = "One way ANOVA with lettering", caption = "By Muhammad Yasir Qurashi" )+
    theme(
    axis.title = element_text(size = 12, color = "black", face = "bold"),
    axis.text = element_text(size = 10, color = "black", face = "bold"), 
    axis.line = element_line(linewidth = 1, color = "black"),
    plot.title = element_text(size = 16, color = "blue", face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12,face = "bold"),
    legend.background = element_rect(color = "black", fill = "grey"),
    legend.position = "right",
    legend.text = element_text(size = 10, face = "bold"),
     panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank()
  ) +
  scale_fill_bmj();p4
```
## Combine Plot

```{r, fig.height = 12, fig.width = 14}
library(patchwork)

finalplot <- (p1 | p2 ) / (p3 | p4) +
  plot_annotation(tag_levels = "A");finalplot
```

Best Regards,

*Muhammad Yasir Qurashi*

Research Data Analysis Tools Mentor