---
title: "Day_06_scientific_visualization_training"
author: "By Muhammad Yasir Qurashi"
date: "2026-03-11"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# **Time Series & Trend Analysis in R**

Time series visualization shows how a variable changes across time.

In research, this helps answer;

* Are values increasing or decreasing over time?

* Is there seasonality?

* Are there anomalies or spikes?

* Are variables moving together or independently?

## Loading dataset

```{r}
data("airquality")
head(airquality)

airquality
```

| Variable | Description                     |
| -------- | ------------------------------- |
| Ozone    | Ozone concentration (ppb)       |
| Solar.R  | Solar radiation (Langley units) |
| Wind     | Wind speed (mph)                |
| Temp     | Temperature (°F)                |
| Month    | Month of observation            |
| Day      | Day of observation              |


## Today's Scientific Questions

* Does ozone pollution increase during summer months?

* Is there a trend between temperature and ozone concentration?

* Do wind patterns reduce pollution levels over time?


## Loading libraries

```{r}
library(tidyverse)
library(RColorBrewer)
library(ggpubr)
library(reshape2)
```


## Data manipulation 


```{r}
# Creating date column in our dateaset that helps us in time series analysis visualization
airquality <-  airquality %>% 
  drop_na() %>% 
  mutate( date = make_date(1973,Month, Day)
    ) %>% 
  arrange(date);airquality

```

## Temporal trend of Ozone Conc.


```{r, fig.width=8}
p1 <- ggplot(airquality, mapping = aes(x = date, y = Ozone)) +
  geom_point(alpha = 0.7, color = "blue", size = 3) +
  geom_smooth( method = "loess", span = 0.3, se = T, color = "#D55E00",
              fill = "#F4A582", linewidth = 1.2) +
  theme_classic() +
    labs( x = "Date (1973)", y = "Ozone Concentration (ppb)", title = "Temporal Trend of Ozone Concentration",  caption = "Made by Muhammad Yasir Qurashi");p1
```

## Trend of Ozone concentration

```{r, fig.width=8}
p2 <- ggplot(airquality, mapping = aes(x = date, y = Ozone)) +
  geom_point(alpha = 1, fill = "black", size = 3) +
  geom_line(color = "blue") +
  theme_classic() +
    labs( x = "Date (1973)", y = "Ozone Concentration (ppb)", title = "Trend of Ozone Concentration",  caption = "Made by Muhammad Yasir Qurashi");p2
```

## Ozone variation across months

```{r, fig.width=8}
p3 <- ggplot(airquality, mapping = aes(x = Day, y = Ozone)) +
  geom_point(alpha = 1, fill = "black", size = 3) +
  geom_line(color = "blue") +
  facet_wrap(~Month) +
  theme_classic() +
    labs( x = "Date (1973)", y = "Ozone Concentration (ppb)", title = "Daily Ozone variation across Months",  caption = "Made by Muhammad Yasir Qurashi");p3
```

## Seasonal distribution of Ozone


```{r}
p4 <- ggplot(airquality, aes(factor(Month), Ozone)) +
  geom_boxplot(fill = "#56B4E9") +
  labs( x = "Month", y = "Ozone (ppb)", title = "Seasonal Distribution of Ozone", caption = "Made by Muhammad Yasir Qurashi") +
  theme_classic();p4
```

## Data conversion into long_format

```{r}

air_long <- airquality %>%
  select(date, Ozone, Temp, Wind) %>%
  pivot_longer(-date)
print(air_long)
```

## Time comparison of Environmental variables

```{r, fig.width=8}
p5 <- ggplot(air_long, aes(date, value, color = name)) +
  geom_line(linewidth = 1) +
  labs( x = "Date", y = "Value", color = "Variable", title = "Time Comparison of Environmental Variables", caption = "Made by Muhammad Yasir Qurashi") +
  theme_bw();p5
```


## Interpretation 

Ozone concentrations exhibited a clear temporal trend across the observation period, with peak levels occurring during mid-summer months. The LOESS-smoothed curve indicated a gradual increase in ozone concentrations from May to July followed by a decline toward September.Seasonal comparison revealed significantly higher ozone concentrations during July compared with early summer months


## Multipanel figure

```{r,fig.height=6, fig.width=12}
library(patchwork)
final_plot_day_06 <- ( p1 | p2 ) / ( p4 | p5 ) +
  plot_annotation(tag_levels = "A");final_plot_day_06
```




Best Regards,

*Muhammad Yasir Qurashi*

Research Data Analysis Tools Mentor


